\thornado-hydro: Generalizing discontinuous galerkin methods for a nuclear equation of state for supernova hydrodynamics

### 2019-02-19 paper/ms.tex
Finished draft of DG subsection

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\todo{Use this form of the Euler Equations?}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details]{leveque:2002} in cartesian coordinates are given by
the equations of conservation of mass,
\beq
  \pderiv{\rho}{t} + \pderiv{}{x}(\rho u) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
	\pderiv{(\rho u)}{t} + \pderiv{}{x}(\rho u^2 + p) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \pderiv{E}{t} + \pderiv{}{x}((E+p)u) = 0
  \label{eq:energyConservation}
\eeq
and conservation of \textcolor{red}{electron density}
\beq
  \pderiv{D}{t} + \pderiv{}{x}(D u) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the fluid
pressure, $D=\rho y_e$ is the \textcolor{red}{what is this?},
and $E=e+\frac{1}{2}\rho u^2$ the total energy (internal plus kinetic).
The system of equations is closed by the EOS which links the pressure to
the quantities evolved by Equations
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
is a more convenient way:

\beq
  \pderiv{\mathbf{U}}{t}
  +\pderiv{\mathbf{F}(\mathbf{U})}{x}
  =0.
\eeq
where $\mathbf{U} =(\rho,\rho u,E, D)^{T}$ is the vector of conserved quantities
and $\mathbf{F}(\mathbf{U})=(\rho u,\rho u^{2}+p,(E+p)u, uD)^{T}$
is the flux vector.

@@

\beq
  \partial_{t} \mathbf{U} + \partial_{x} \mathbf{F}(\mathbf{U}) = 0
\eeq

We can rewrite this further:

\beq
  \pderiv{\mathbf{U}}{t}
  + \pderiv{\mathbf{U}}{x} \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}}
  = 0
  \label{charEq}
\eeq
Because the Euler equations for a systemer of hyperblic
partial differential equations \citep[see, e.g.,][]{leveque:1992}, we can decompose the
Jacobian of the flux vector as

\beq
  \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}} =
  R \Lambda R^{-1}
\eeq
where the columns of $R$ contain the right eigenvectors of the Jacobian,
the rows of $R^{-1}$ contain the left eigenvectors, and
$\Lambda$ is a diagonal matrix containing the eigenvalues of the Jacobian.
At this point, we will introduce the characteristic variable
$\mathbf{w} = R^{-1}\mathbf{U}$. Multiplying both sides of Equation
\eqref{charEq} by $R^{-1}$, we are able to linearize the system of equations to
a system of advection equations

>>>>

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\todo{Use this form of the Euler Equations?}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details]{leveque:2002} in cartesian coordinates are given by
the equations of conservation of mass,
\beq
  \pderiv{\rho}{t} + \pderiv{}{x}(\rho u) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
	\pderiv{(\rho u)}{t} + \pderiv{}{x}(\rho u^2 + p) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \pderiv{E}{t} + \pderiv{}{x}((E+p)u) = 0
  \label{eq:energyConservation}
\eeq
and conservation of \textcolor{red}{electron density}
\beq
  \pderiv{D}{t} + \pderiv{}{x}(D u) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the fluid
pressure, $D=\rho y_e$ is the \textcolor{red}{what is this?},
and $E=e+\frac{1}{2}\rho u^2$ the total energy (internal plus kinetic).
The system of equations is closed by the EOS which links the pressure to
the quantities evolved by Equations
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
in a more convenient way:

\beq
  \pderiv{\mathbf{U}}{t}
  +\pderiv{\mathbf{F}(\mathbf{U})}{x}
  =0,
\eeq
where $\mathbf{U} =(\rho,\rho u,E, D)^{T}$ is the vector of conserved quantities
and $\mathbf{F}(\mathbf{U})=(\rho u,\rho u^{2}+p,(E+p)u, uD)^{T}$
is the flux vector.

@@

\beq
  \partial_{t} \mathbf{U} + \partial_{x} \mathbf{F}(\mathbf{U}) = 0
  \label{eq:ode}
\eeq
Where $\mathbf{U}$ is the evolved state vector and $\mathbf{F}(\mathbf{U})$ is
the flux. In order to solve Equation \eqref{eq:ode}, we divide the computational
domain $D\subset \mathbb{R}$ into a disjoint union $\mathcal{T}$ of open elements
$\bK$ such that $D = \cup_{\bK \in \cT}\bK$. Each element $\bK$ is a box in the
coordinates
\beq
  \bK=\{\,\vect{x} = (\xL,x_H) : x \in K \},
\eeq
We let the approximation space $\mathbb{V}^{k}$ for the DG method
be constructed from the tensor product of one-dimensional
polynomials of degree $k$. Note that functions in $\mathbb{V}^{k}$
can be discontinuous across element interfaces. The DG problem is then to find
$\mathbf{U}_h \in \mathbb{V}^{k}$ which approximates $\mathbf{U}$ in Equation
\eqref{eq:ode}, such that $\forall$ $\phi \in \mathbb{V}^{k}$ and $\bK \in \cT$

\beq
  \partial_{t} \int_{\bK}\mathbf{U}_h\, \phi\, dx +
  \int_{\bK}\partial_{x}\, \mathbf{F(\mathbf{U})}\, \phi\, dx = 0.
\label{eq:semidiscreteDG_almost}
\eeq
Integrating the second term by parts, this becomes
\beq
\begin{split}
  & \partial_{t} \int_{\bK}\mathbf{U_{h}}\, \phi\, dx +
   \widehat{\mathbf{F}}(\mathbf{U}_h)\, \phi^{-} \big|_{x_{H}} \\
  & + \widehat{\mathbf{F}}(\mathbf{U}_h)\, \phi^{+} \big|_{x_{L}} +
   \int_{\bK}\mathbf{F(\mathbf{U}_h)}\, \partial_{x}\phi\, dx = 0.
\label{eq:semidiscreteDG}
\end{split}
\eeq
In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\vect{F}}(\vect{U}_{h})$ is a
numerical flux approximating the flux on the boundary of $\bK$.
The numerical flux function is evaluated using values from
both boundaries of an element; i.e.,
\begin{equation}
  \widehat{\vect{F}}(\vect{U}_{h})=\vect{f}(\vect{U}_{h}(x^{-},\tilde{\bx}),\vect{U}_{h}(x^{+},\tilde{\bx})),
\end{equation}
where superscripts $-/+$
indicate that the function is evaluated to the immediate left/right of the
interface. We use the Harten-Lax-van Leer (HLL) flux \citep{harten:1983} or
the HLLC flux \citep{toro:1994,mignone:2005} for all the numerical
experiments presented in Section \ref{sec:tests}.

In each element $K$, we use a nodal representation of the conserved variables
$\mathbf{U}$:

\beq
\begin{split}
  \mathbf{U}(x,t) \approx\, & \mathbf{U}_h (x,t) =
  \sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad \\
  & \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{split}
\eeq
are Lagrange polynomials defined on $I = \{ \eta : \eta \in (-0.5,0.5) \}$,
and are constructed to interpolate the node set
$S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$. The spatial coordinate $x$ and the
reference coordinate $\eta$ are related by the
mapping $x(\eta)=\xL+(0.5+\eta)\,\Delta x$.
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$,
so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$
with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights
$\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$ in order to
evaluate the integrals in Eq.~\eqref{eq:semidiscreteDG}..
We use the $M$-point Legendre-Gauss quadrature, which is exact for
polynomials of degree $\le 2M-1$.
Then, if $P_{h}(x)$ is such a polynomial, we have
\beq
  \f{1}{\Delta x}\int_{K}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).
\eeq

For the sake of efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a
spectral-type nodal collocation DG approximation \citep{bassi:2013}.
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG},
letting $\phi(x)=\ell_{k}(x)$, using the quadratures defined above,
we obtain\footnote{Are these summations?}
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,\phi\,dx
  &\approx w_{k}\,\pd{}{t}\vect{U}_{k}\,\Delta x
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative piece.
Similarly, the last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}
becomes
\beq
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{\phi}{x}\,dx
  \approx \sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\eeq
Now we may combine Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in
Eq.~\eqref{eq:semidiscreteDG} resulting in the semi-discrete form
\beq
\begin{split}
  \deriv{\vect{U}_{k}}{t}
  & =-\f{1}{w_{k}\Delta x}
  \Big\{
  \Big[\,
    \widehat{\vect{F}}\ell_{k}\big|_{x_{H}}
     -\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big] \\
  & -\sum_{q=1}^{N}w_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\}.
  \label{eq:semidiscreteDiscretized}
\end{split}
\eeq
We now have a system of ordinary
differential equations (ODEs), which may be evolved in time with an ODE solver.
In Section~\ref{sec:tests} we use the third-order strong
stability-preserving Runge-Kutta (SSP-RK3) method \citep{shu:1988}.

\subsection{Slope Limiting}

\subsection{Characteristic Decomposition}
Experience has shown...
We can rewrite this further:

\beq
  \pderiv{\mathbf{U}}{t}
  + \pderiv{\mathbf{U}}{x} \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}}
  = 0
  \label{eq:charEq}
\eeq
Because the Euler equations for a systemer of hyperblic
partial differential equations \citep[see, e.g.,][]{leveque:1992}, we can decompose the
Jacobian of the flux vector as

\beq
  \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}} =
  R \Lambda R^{-1}
\eeq
where the columns of $R$ contain the right eigenvectors of the Jacobian,
the rows of $R^{-1}$ contain the left eigenvectors, and
$\Lambda$ is a diagonal matrix containing the eigenvalues of the Jacobian.
At this point, we will introduce the characteristic variable
$\mathbf{w} = R^{-1}\mathbf{U}$. Multiplying both sides of Equation
\eqref{eq:charEq} by $R^{-1}$, we are able to linearize the system of equations to
a system of advection equations
